---
title: "Overview of lineplots"
author: "Anna Quaglieri"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      collapse = TRUE,
                      comment = "#>")
```


# Scope of this package

The `lineplots` package was developed to standardise the variant calling output from a cohort of patients for a set of genes of interest to then produce plots over time. It has specific functions built to interact with the output from `superFreq` (`extract_goi_superfreq()`) as well as more generic import functions to integrate calls from various callers (`import_any`) or clinical information (`import_clinical`). It also provide a parsing function, `parse_vcf_output()` which takes a `VCF` file as input and outputs a parsed `tibble` with standardised names for VCF fields. Currently, `parse_vcf_output()` supports ouputs from the following callers run in tumour-only mode: `GATK3 MuTect2`, `VarScan2`, `VarDict`. It can also enable parsing of the annotation field added by the [Variant Effect Predictor](https://asia.ensembl.org/info/docs/tools/vep/index.html). The package was developed with variants from RNA-Seq samples but there would be in principle no difference if variants were called using DNA-Seq. 
It was developed as part of the work done on an RNA-Seq cohort of Leukemia patients, of relatively small sample size (~30 patients), with multiple samples at different time points, sequenced across different batches and tissues. In these type of studies it is often the case that clinicians or biologists needs to explore the information for each patient, for a particular set of genes of interest. Time-series lineplots over time as well as heatmaps showing the shifts in variant allele frequency (VAF) and tumour content help with exploring the various genetic results generated from sequencing data. A Shiny app, [`MuExplore`](https://github.com/annaquaglieri16/mutexplore), is also under development to allow interactive exploration of the standardised data across all the patients. 

## A multidude of outputs

Calling variants from RNA-Seq samples is increasing in popularity due to the large number of RNA-Seq data produced **[papers]**. Several benchmarking papers have been published comparing the performance of different callers [] and in recent years several new methods have been developed specifically to improve INDEL detection in RNA-Seq samples []. As part of a project done on two Leukemia clinical trials we defined a pipeline to call variants in RNA-Seq (https://rna-mutation-calls.netlify.com/) and use [superFreq](https://github.com/ChristofferFlensburg/superFreq) to analyse clonal tracking and mutational development across treatment. We used combined calls from the [VarDict](https://github.com/AstraZeneca-NGS/VarDict) [@Lai2016-ws] and [km](https://github.com/iric-soft/km/tree/master/km/tools#output-description-1) [@Software-km] algorithms to define INDEL calls.  

After calling SNVs, CNVs and INDELs with different programs, it is often useful to combine all the results in a standardised way. Usually different programs output variants in different ways using slightly different conventions which makes it challenging to easily combine the results. For example, superFreq returns SNVs, CNVs and clone tracking in `Rdata` files as well as in `csv` files; `VarDict` uses the standard `VCF` output but with different specification of quality measures compared to, for example, `MuTect2` or `VarScan2`; `km` outputs INDELs in tab delimited format. This makes it challenging to both create a comprehensive overview of the results for one sample as well as to compare results across callers. Another challenge comes from calling variants independently at different time points for the same patient. Variants might be present before treatment but absent after treatment and when a patient relapses or vice versa. This required to fill in the missing time points especially for visualisations purposes. While `superFreq` takes this problem into account by keeping any variants found at different time points, other programs like `km` or `VarDict` don’t allow that. 

# Setup 

The `lineplots` R package comprises several helper and import functions that can be used to convert output from any of the callers mentioned above to a standardised format.

```{r eval=FALSE}
library(devtools)
devtools::install_github("annaquaglieri16/lineplots")
```


```{r eval=TRUE}
library(lineplots)
```

```{r}
ls("package:lineplots")
```

# From `VCF` to `tibble`: `parse_vcf_output()`

The function `parse_vcf_output()` allows to convert a `.vcf` file generated by three different callers `GATK3 MuTect2`, `VarDict` or `VarScan`, to a data frame with standardised fields. This is because different callers annotate the `INFO` field in the `VCF` ouput in different ways, using different names for read depth, variant allele frequency etc... The output from `parse_vcf_output()` can then be used for `import_any_for_lineplots()` to combine variants and clinical information and fill in missing time points. If the variants in the `VCF` file were also annotated with the Variant Effect Predictor (VEP) [@McLaren2016-lv], the one can set `vep = TRUE` to parse the extra fields. 

It is to be noted that VEP can annotate the same variant multiple times, depending on whether a variants falls on several transcripts. VEP pastes together different annotations on the same line. `parse_vcf_output` will reshape the input variants and return it in a long format by stacking all the annotations one underneath each other. This has the potential of largely increasing the size of the ouput file and it is suggested to subset variants based on genes of interest, if the input `VCF` file is too large.   

The `parse_vcf_output` functions is built on top of the [`VariantAnnotation`](https://bioconductor.org/packages/release/bioc/html/VariantAnnotation.html) Bioconductor package [@Obenchain2014-cd] to quickly read large `VCF` files into `R` and extract fields.

The example `VCF` files below were annotated with VEP. Also `germline_varscan.vcf` and `germline_vardict.vcf` are available as example. 

```{r include=TRUE,echo=TRUE}
annot_vcf_mutect <- system.file("extdata", "germline_mutect.vcf", package = "lineplots")
annot_vcf_mutect
```

`VCF` files from `VarDict` and `VarScan` can be parsed by specifying `caller = "vardict"` or `caller = "varscan"`.

```{r include=FALSE,echo=TRUE}
parsed_vcf_mutect <- lineplots::parse_vcf_output(annot_vcf_mutect, caller = "mutect",sample_name = "Sample1",vep = TRUE)
```

```{r parsed-vcf-mut}
knitr::kable(parsed_vcf_mutect[1:10,],caption = "Parsed MuTect2 output, from VCF to data frame.")
```

You can find in [Standardise VCFs output across callers](#st-variants) more details about the exact fields that I extract from a caller's `VCF` file to populate the standardised fields given in ouput.

# `import` functions

In general, the import functions can take in input variants called by a caller as well as clinical information for all the patients in the cohort under investigation. Every input function requires a `patientID` to be provided and a set of `studyGenes` to restrict the analysis on. `studygenes` are expected to be a character vector of NCBI gene `Symbols`. This means that variants need to be annotated with the gene they belong to before `lineplots` can be used. The clinical data at every time point is needed because if a variant is not detected at one time point for a patient, then the input variants will be extended to include all time points initially available for that patient but with default values for missing fields (e.g. VAF = 0). This is useful for visualisation purposes where one needs to track the development of mutation over time.

Figure XX is a graphic summary of the input required for every `import` function.

**add pic with input and output of every function**
**Picture of filling gaps strategy**


# Standardise VCFs output across callers {#st-variants}

Following, is the list of fields which are given as output after parsing a VCF file with `parse_vcf_output()`. Here is a description of what information was extracted for every caller to populate each standardised fields:

* *Location*: `CHR_POS` where `CHR` and `POS` are standard `VCF` fields for every caller.
* *caller*: the caller passed to the `caller` argument.
* *chrom*: field `CHROM` in the `VCF` file.
* *pos*: field `POS` in the `VCF` file.
* *ref*: field `REF` in the `VCF` file.
* *alt*: field `ALT` in the `VCF` file.
* *qual*: This field is not reported consistently from the VCF output of the three callers but I made sure that its meaning is consistent and that it represents the average base quality at that position. Below is a description of how it is extracted from each caller.
    - In `MuTect2` the `QSS` field in the `FORMAT` fields reports the sum of the base qualities for the reference and alternative alleles separated by a comma. We used the reference and alternative depths at each position to compute the overall average base quality at that position. This quantity will populate the final `qual` field. 
    - In `VarScan` this field is the average of the `RBQ` and `ABQ` fields from the `FORMAT` fields. They are defined respectively as the average quality of reference and alternative supporting bases in the header of the `VCF` file.
    - In `VarDict` there are two `QUAL` fields, one is the standard 6th field of a `VCF` file and the other one is reported in the `INFO` fields. We will use the latter to populate the `qual` field in our analysis since it is the one defined as the average base quality at a variant position in the header of the `VCF`. `VarDict` uses a threshold of `QUAL >= 25` to report a variant. 

* *filter*:  standard field `FILTER` in the `VCF` file. Each caller populates this field in different ways depending on the characteristics of the algorithm. In general, the entries for this field can be either `PASS` if that mutation passes all the filters defined by a caller or a description of the reason for filtering. The possible descriptions and their meaning can be found in the header of the VCF file generated by the caller.

* *genotype*: standard `GT` field in the `FORMAT` fields of the `VCF` file.

* *tot_depth*: total read depth at each position as estimated by the caller. This information can be reported differently by each caller. `VarDict` and `VarScan`  record it in the `DP` field while `MuTect2` records the reference and alternative depth in the `AD` columns and their sum was used to define the `total_depth`.

* *VAF*: variant allele frequency for the variants recorded at that position. `VarScan` records it in the `FREQ` field while `MuTect` and `VarDict` in the `AF` field. 

* *ADJVAF_ADJ_indels*: field `ADJAF` only reported by `VarDict` and it represents the adjusted variant allele frequency for indels due to local realignment.

* *ref_depth*, *alt_depth*, *ref_forw*, *ref_rev*, *alt_forw* and *alt_rev*: these fields represent the breakdown of supporting reference/alternative and forward/reverse reads at each location. Below I described what fields I used from every caller to extract these values. The fields are listed in order:
    - `MuTect2`: *ref_depth* and *alt_depth* are the comma separated values reported in the field `AD`; *ref_forw*, *ref_rev*, *alt_forw* and *alt_rev* are respectively the `MuTect2` fields `REF_F1R2`, `REF_F2R1`, `ALT_F1R2` and `ALT_F2R1`. 
    - `VarScan2`: in order the features listed above are extracted from the the fields `RD`, `AD`, `RDF`, `RDR`, `ADF` and `ADR`.
    - `VarDict`: *ref_depth* and *alt_depth* are the fields `REF` and `ALT` in the VCF file; the field `REFBIAS` contains comma separated values representing  *ref_forw* and *ref_rev* and `VARBIAS` contains comma separated values representing *alt_forw* and *alt_rev*.
  
If the `VCF` file was also annotated using the Variant Effect Predictor (VEP) [@McLaren2016-lv]  
All the other fields are exactly as generated by `VEP`: `Allele`, `Consequence`, `IMPACT`, `SYMBOL`, `Gene`, `Feature_type`, `Feature`, `BIOTYPE`, `EXON`, `INTRON`, `HGVSc`,  `HGVSp`, `cDNA_position`, `CDS_position`, `Protein_position`   `Amino_acids`, `Codons`, `Existing_variation` `DISTANCE`, `STRAND`, `FLAGS`, `VARIANT_CLASS`, `SYMBOL_SOURCE`, `HGNC_ID`, `CANONICAL`, `TSL`, `APPRIS`, `CCDS`, `ENSP`, `SWISSPROT`, `TREMBL`, `UNIPARC`, `GENE_PHENO`, `SIFT`, `PolyPhen`, `DOMAINS`, `AF`, `AFR_AF`, `AMR_AF`, `EAS_AF`, `EUR_AF`, `SAS_AF`, `AA_AF`,  `EA_AF`, `ExAC_AF`, `ExAC_Adj_AF`, `ExAC_AFR_AF`, `ExAC_AMR_AF`, `ExAC_EAS_AF`, `ExAC_FIN_AF`, `ExAC_NFE_AF`, `ExAC_OTH_AF`,  `ExAC_SAS_AF`, `MAX_AF`, `MAX_AF_POPS`, `CLIN_SIG`, `SOMATIC`, `PHENO`, `PUBMED`,`MOTIF_NAME`, `MOTIF_POS`, `HIGH_INF_POS`, `MOTIF_SCORE_CHANGE` `SampleName`, `IMPACT_rank`. 
Visit the `VEP` page to find more information https://asia.ensembl.org/info/docs/tools/vep/index.html. 
